---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 1"
author: "Doris Mbitazi Asongafac"
date: "2026-05-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth = 80)
```

## Load required libraries and datasets
Note that you will need to install these libraries if you have never used these before.

```{r load-libraries, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(readr)
library(stringr)
# library(ggmosaic) # not available in this environment
```

```{r load-data}
filePath <- "C:/Users/doris/Downloads/Quantium Analytics/"
transactionData <- fread(paste0(filePath, "QVI_transaction_data.csv"))
customerData    <- fread(paste0(filePath, "QVI_purchase_behaviour.csv"))
```

## Exploratory data analysis

The first step in any analysis is to first understand the data. Let's take a look at each of the datasets provided.

### Examining transaction data

We can use `str()` to look at the format of each column and see a sample of the data. Let's check if columns we would expect to be numeric are in numeric form and date columns are in date format.

```{r examine-transaction-data}
# Examine transaction data
str(transactionData)
```

We can see that the date column is in an integer format. Let's change this to a date format.

```{r convert-date}
# Convert DATE column to a date format
# CSV and Excel integer dates begin on 30 Dec 1899
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")

# Verify the conversion
str(transactionData)
```

We should check that we are looking at the right products by examining PROD_NAME.

```{r examine-prod-name}
# Examine PROD_NAME
transactionData[, .N, PROD_NAME]
```

Looks like we are definitely looking at potato chips but how can we check that these are all chips? We can do some basic text analysis by summarizing the individual words in the product name.

```{r analyse-product-words}
# Examine the words in PROD_NAME to see if there are any incorrect entries
# such as products that are not chips
productWords <- data.table(unlist(strsplit(unique(transactionData[, PROD_NAME]), " ")))
setnames(productWords, "words")

# Removing digits
productWords <- productWords[grepl("\\d", words) == FALSE, ]

# Removing special characters
productWords <- productWords[grepl("[[:alpha:]]", words), ]

# Sort distinct words by frequency, highest to lowest
productWords[, .N, words][order(N, decreasing = TRUE)]
```

There are salsa products in the dataset but we are only interested in the chips category, so let's remove these.

```{r remove-salsa}
# Remove salsa products
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]
transactionData <- transactionData[SALSA == FALSE, ][, SALSA := NULL]
```

Next, we can use `summary()` to check summary statistics such as mean, min and max values for each feature to see if there are any obvious outliers in the data and if there are any nulls in any of the columns.

```{r initial-summary}
# Summarise the data to check for nulls and possible outliers
summary(transactionData)
```

There are no nulls in the columns but product quantity appears to have an outlier which we should investigate further. Let's investigate the case where 200 packets of chips are bought in one transaction.

```{r find-outlier}
# Filter the dataset to find the outlier
transactionData[PROD_QTY == 200, ]
```

There are two transactions where 200 packets of chips are bought in one transaction and both of these transactions were by the same customer.

```{r check-outlier-customer}
# Let's see if the customer has had other transactions
transactionData[LYLTY_CARD_NBR == 226000, ]
```

It looks like this customer has only had the two transactions over the year and is not an ordinary retail customer. The customer might be buying chips for commercial purposes instead. We'll remove this loyalty card number from further analysis.

```{r remove-outlier-customer}
# Filter out the customer based on the loyalty card number
transactionData <- transactionData[LYLTY_CARD_NBR != 226000, ]

# Re-examine transaction data
summary(transactionData)
```

That's better. Now, let's look at the number of transaction lines over time to see if there are any obvious data issues such as missing data.

```{r count-transactions-by-date}
# Count the number of transactions by date
transactionData[, .N, by = DATE]
```

There's only 364 rows, meaning only 364 dates which indicates a missing date. Let's create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a chart of number of transactions over time to find the missing date.

```{r transactions-over-time, fig.align="center"}
# Create a sequence of dates and join this to the count of transactions by date
allDates <- data.table(seq(as.Date("2018/07/01"), as.Date("2019/06/30"), by = "day"))
setnames(allDates, "DATE")
transactions_by_day <- merge(allDates, transactionData[, .N, by = DATE], all.x = TRUE)

# Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

# Plot transactions over time
ggplot(transactions_by_day, aes(x = DATE, y = N)) +
  geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that there is an increase in purchases in December and a break in late December. Let's zoom in on this.

```{r zoom-december, fig.align="center"}
# Filter to December and look at individual days
ggplot(transactions_by_day[month(DATE) == 12, ], aes(x = DATE, y = N)) +
  geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 day") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that the increase in sales occurs in the lead-up to Christmas and that there are zero sales on Christmas day itself. This is due to shops being closed on Christmas day.

Now that we are satisfied that the data no longer has outliers, we can move on to creating other features such as brand of chips or pack size from PROD_NAME. We will start with pack size.

```{r create-pack-size}
# Pack size
# We can work this out by taking the digits that are in PROD_NAME
transactionData[, PACK_SIZE := parse_number(PROD_NAME)]

# Always check your output
# Let's check if the pack sizes look sensible
transactionData[, .N, PACK_SIZE][order(PACK_SIZE)]
```

The largest size is 380g and the smallest size is 70g - seems sensible!

```{r plot-pack-size}
# Plot a histogram of PACK_SIZE
# It is a categorical variable even though it is numeric
hist(transactionData[, PACK_SIZE])
```

Pack sizes created look reasonable and now to create brands, we can use the first word in PROD_NAME to work out the brand name.

```{r create-brand-name}
# Brands
transactionData[, BRAND := toupper(substr(PROD_NAME, 1, regexpr(pattern = " ", PROD_NAME) - 1))]

# Checking brands
transactionData[, .N, by = BRAND][order(-N)]
```

Some of the brand names look like they are of the same brands - such as RED and RRD, which are both Red Rock Deli chips. Let's combine these together.

```{r clean-brand-names}
# Clean brand names
transactionData[BRAND == "RED",    BRAND := "RRD"]
transactionData[BRAND == "SNBTS",  BRAND := "SUNBITES"]
transactionData[BRAND == "INFZNS", BRAND := "INFUZIONS"]
transactionData[BRAND == "WW",     BRAND := "WOOLWORTHS"]
transactionData[BRAND == "SMITH",  BRAND := "SMITHS"]
transactionData[BRAND == "NCC",    BRAND := "NATURAL"]
transactionData[BRAND == "DORITO", BRAND := "DORITOS"]
transactionData[BRAND == "GRAIN",  BRAND := "GRNWVES"]

# Check again
transactionData[, .N, by = BRAND][order(BRAND)]
```

### Examining customer data

Now that we are happy with the transaction dataset, let's have a look at the customer dataset.

```{r examine-customer-data}
# Examining customer data
str(customerData)
summary(customerData)
```

```{r customer-distributions}
# Examining the values of lifestage and premium_customer
customerData[, .N, by = LIFESTAGE][order(-N)]
customerData[, .N, by = PREMIUM_CUSTOMER][order(-N)]
```

```{r merge-data}
# Merge transaction data to customer data
data <- merge(transactionData, customerData, all.x = TRUE)
```

As the number of rows in `data` is the same as that of `transactionData`, we can be sure that no duplicates were created. This is because we created `data` by setting `all.x = TRUE` (a left join).

Let's also check if some customers were not matched on by checking for nulls.

```{r check-nulls}
data[is.null(LIFESTAGE), .N]
data[is.null(PREMIUM_CUSTOMER), .N]
```

Great, there are no nulls! So all our customers in the transaction data have been accounted for in the customer dataset.

```{r save-data}
fwrite(data, paste0(filePath, "QVI_data.csv"))
```

Data exploration is now complete!

## Data analysis on customer segments

Now that the data is ready for analysis, we can define some metrics of interest to the client:

- Who spends the most on chips (total sales), describing customers by lifestage and how premium their general purchasing behavior is
- How many customers are in each segment
- How many chips are bought per customer by segment
- What's the average chip price by customer segment

Let's start with calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER and plotting the split by these segments to describe which customer segment contribute most to chip sales.

```{r total-sales-by-segment, fig.width=10, fig.align="center"}
# Total sales by LIFESTAGE and PREMIUM_CUSTOMER
sales <- data[, .(SALES = sum(TOT_SALES)), .(LIFESTAGE, PREMIUM_CUSTOMER)]

# Create plot
ggplot(sales, aes(x = LIFESTAGE, y = SALES, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "Lifestage", y = "Total Sales ($)",
    title = "Total Chip Sales by Customer Segment",
    fill = "Customer Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Sales are coming mainly from Budget - older families, Mainstream - young singles/couples, and Mainstream - retirees.

Let's see if the higher sales are due to there being more customers who buy chips.

```{r customers-by-segment, fig.width=10, fig.align="center"}
# Number of customers by LIFESTAGE and PREMIUM_CUSTOMER
customers <- data[, .(CUSTOMERS = uniqueN(LYLTY_CARD_NBR)), .(LIFESTAGE, PREMIUM_CUSTOMER)][order(-CUSTOMERS)]

# Create plot
ggplot(customers, aes(x = LIFESTAGE, y = CUSTOMERS, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "Lifestage", y = "Number of Customers",
    title = "Number of Chip-Buying Customers by Segment",
    fill = "Customer Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

There are more Mainstream - young singles/couples and Mainstream - retirees who buy chips. This contributes to there being more sales to these customer segments but this is not a major driver for the Budget - Older families segment.

Higher sales may also be driven by more units of chips being bought per customer. Let's have a look at this next.

```{r avg-units-per-customer, fig.width=10, fig.align="center"}
# Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER
avg_units <- data[, .(AVG = sum(PROD_QTY) / uniqueN(LYLTY_CARD_NBR)),
                  .(LIFESTAGE, PREMIUM_CUSTOMER)][order(-AVG)]

# Create plot
ggplot(avg_units, aes(weight = AVG, x = LIFESTAGE, fill = PREMIUM_CUSTOMER)) +
  geom_bar(position = position_dodge()) +
  labs(
    x = "Lifestage", y = "Avg units per transaction",
    title = "Units per customer"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Older families and young families in general buy more chips per customer.

Let's also investigate the average price per unit chips bought for each customer segment as this is also a driver of total sales.

```{r avg-price-per-unit, fig.width=10, fig.align="center"}
# Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER
avg_price <- data[, .(AVG = sum(TOT_SALES) / sum(PROD_QTY)),
                  .(LIFESTAGE, PREMIUM_CUSTOMER)][order(-AVG)]

# Create plot
ggplot(avg_price, aes(weight = AVG, x = LIFESTAGE, fill = PREMIUM_CUSTOMER)) +
  geom_bar(position = position_dodge()) +
  labs(
    x = "Lifestage", y = "Avg price per unit",
    title = "Price per unit"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Mainstream midage and young singles and couples are more willing to pay more per packet of chips compared to their budget and premium counterparts. As the difference in average price per unit isn't large, we can check if this difference is statistically different.

```{r t-test}
# Perform an independent t-test between mainstream vs premium and budget
# midage and young singles and couples
pricePerUnit <- data[, price := TOT_SALES / PROD_QTY]

t.test(
  data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
         PREMIUM_CUSTOMER == "Mainstream", price],
  data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
         PREMIUM_CUSTOMER != "Mainstream", price],
  alternative = "greater"
)
```

The t-test results in a p-value < 2.2e-16, i.e. the unit price for mainstream, young and mid-age singles and couples are significantly higher than that of budget or premium, young and midage singles and couples.

## Deep dive into specific customer segments for insights

We have found quite a few interesting insights that we can dive deeper into. We might want to target customer segments that contribute the most to sales to retain them or further increase sales. Let's look at Mainstream - young singles/couples. For instance, let's find out if they tend to buy a particular brand of chips.

```{r brand-affinity, fig.align="center"}
# Deep dive into Mainstream, young singles/couples
segment1 <- data[LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream", ]
other    <- data[!(LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream"), ]

# Brand affinity compared to the rest of the population
quantity_segment1 <- segment1[, sum(PROD_QTY)]
quantity_other    <- other[, sum(PROD_QTY)]

quantity_segment1_by_brand <- segment1[, .(targetSegment = sum(PROD_QTY) / quantity_segment1), by = BRAND]
quantity_other_by_brand    <- other[,    .(other = sum(PROD_QTY) / quantity_other),             by = BRAND]

brand_proportions <- merge(quantity_segment1_by_brand, quantity_other_by_brand)[,
  affinityToBrand := targetSegment / other]

brand_proportions[order(-affinityToBrand)]
```

We can see that:

- Mainstream young singles/couples are 23% more likely to purchase Tyrrells chips compared to the rest of the population
- Mainstream young singles/couples are 56% less likely to purchase Burger Rings compared to the rest of the population

Let's also find out if our target segment tends to buy larger packs of chips.

```{r pack-size-affinity, fig.align="center"}
# Preferred pack size compared to the rest of the population
quantity_segment1_by_pack <- segment1[, .(targetSegment = sum(PROD_QTY) / quantity_segment1), by = PACK_SIZE]
quantity_other_by_pack    <- other[,    .(other = sum(PROD_QTY) / quantity_other),             by = PACK_SIZE]

pack_proportions <- merge(quantity_segment1_by_pack, quantity_other_by_pack)[,
  affinityToPack := targetSegment / other]

pack_proportions[order(-affinityToPack)]
```

Mainstream young singles/couples are 27% more likely to purchase a 270g pack of chips compared to the rest of the population. Let's check what brand sells this pack size.

```{r check-270g-brand}
data[PACK_SIZE == 270, unique(PROD_NAME)]
```

Twisties are the only brand offering 270g packs and so this may instead be reflecting a higher likelihood of purchasing Twisties.

## Conclusion

Sales have mainly been driven by Budget - older families, Mainstream - young singles/couples, and Mainstream - retirees shoppers. The high spend in chips for mainstream young singles/couples and retirees is due to there being more of them than other buyers. Mainstream, midage and young singles and couples are also more likely to pay more per packet of chips, which is indicative of impulse buying behavior.

Mainstream young singles and couples are 23% more likely to purchase Tyrrells chips compared to the rest of the population. The Category Manager may want to increase the category's performance by off-locating some Tyrrells and smaller packs of chips in discretionary space near segments where young singles and couples frequent more often to increase visibility and impulse behavior.
